PERs from Projections for Binding-time Analysis

### 2017-04-17 pubs/demand_analysis/jlsc.tex

\documentclass[11pt]{article}

\usepackage{geometry}
\geometry{textwidth=7.5in,
  textheight=9in}

%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkScale{6}
%\SetWatermarkColor{yellow}

\begin{document}

\title{PERs from Partial Projections for Higher-order Polymorphic Demand Analysis}
\author{Kei Davis\\
  Los Alamos National Laboratory\\
  kei at lanl.gov}
\maketitle

@@

@@

\com{\data}{{\it data}_{\stt}}
\com{\for}{{\it for}_{\stt}}
\com{\fac}{{\it fac}_{\stt}}
\com{\unfac}{{\it unfac}_{\stt}}

@@

\com{\tmn}{\theta_{mn}}
\com{\tnm}{\theta_{nm}}
\com{\tmy}{\theta_{m\infty}}
\com{\tyy}{\theta_{\infty\infty}}
\com{\tiy}{\theta_{i\infty}}
\com{\tym}{\theta_{\infty{}m}}
\com{\tyz}{\theta_{\infty{\rm0}}}
\com{\tyo}{\theta_{\infty{\rm1}}}
\com{\tyi}{\theta_{\infty{}i}}
\com{\tyip}{\theta_{\infty{}i+{\rm1}}}
%\com{\th}{\theta}

@@

\com{\A}{\alpha}
\com{\lA}{\lift{\A}}
\com{\sA}{\slift{\A}}
\com{\B}{\beta}
\com{\lB}{\lift{\B}}
%\com{\sB}{\slift{\B}}
\com{\G}{\gamma}
\com{\lG}{\lift{\G}}
%\com{\sG}{\slift{\G}}
\com{\D}{\delta}
\com{\lD}{\lift{\D}}
\com{\sD}{\slift{\D}}
\com{\BOT}{{\it BOT}}
\com{\lBOT}{\lift{\BOT}}
\com{\sBOT}{\slift{\BOT}}
\com{\ID}{{\it ID}}
\com{\LAM}{{\it LAM}}
\com{\lID}{\lift{\ID}}
\com{\sID}{\slift{\ID}}
\com{\ABS}{{\it ABS}}
\com{\STR}{{\it STR}}
\com{\Ab}{\alpha_{@@@b@@@}}
\com{\Ax}{\alpha_{@@@x@@@}}
\com{\Ay}{\alpha_{@@@y@@@}}
\com{\Az}{\alpha_{@@@z@@@}}
\com{\Axs}{\alpha_{@@@xs@@@}}
\com{\Ays}{\alpha_{@@@ys@@@}}
\com{\Azs}{\alpha_{@@@zs@@@}}

@@

\com{\oy}{\overline{@y@}}

\com{\xnot}{{\rm@x@_0}}
\com{\xone}{{\rm@x@_1}}
\com{\xtwo}{{\rm@x@_2}}
\com{\xthree}{{\rm@x@_3}}
\com{\xxi}{{@x@_i}}
\com{\xn}{{@x@_n}}

\com{\ee}{@e@}
\com{\ex}{@x@}
\com{\evar}{@x@}
\com{\enum}{@n@_i}
\com{\esum}{\eone\ @+@\ \etwo}
\com{\elam}{@\x.e@}
\com{\ehlam}{@\#x.e@}
\com{\elamt}{@\x:T.e@}
\com{\ehlamt}{@\#x:T.e@}
\com{\eapp}{\eone~\etwo}
\com{\ehapp}{@app# @\eone~\etwo}
\com{\efapp}{@f e@}
\com{\eunit}{@()@}
\com{\etup}{@(@\eone@,@\ldots@,@\en@)@}
\com{\esel}{@let (@\xone@,@\ldots@,@\xn@) =@\ \enot\ @in@\ \eone}
\com{\econ}{@c@_i\ @e@}
\com{\ecase}{@case@\ \enot\ @of @\cone\ \xone\ @->@\ \eone@;@\ \ldots@;@\ \cn\ \xn\ @->@\ \en}
\com{\efix}{@fix e@}
\com{\ehfix}{@fix# e@}
\com{\estr}{@seq@~\eone~\etwo}

\com{\enot}{{\rm@e@_0}}
\com{\eone}{{\rm@e@_1}}
\com{\etwo}{{\rm@e@_2}}
\com{\ei}{{@e@_i}}
\com{\en}{{@e@_n}}

\com{\senot}{{\rm@@@e@@@_{\subtypesubstyle0}}}
\com{\seone}{{\rm@@@e@@@_{\subtypesubstyle1}}}
\com{\setwo}{{\rm@@@e@@@_{\subtypesubstyle2}}}
\com{\sei}{{@@@e@@@_{\subtypesubstyle{}i}}}
\com{\sen}{{@@@e@@@_{\subtypesubstyle{}n}}}
\com{\setup}{{@@@(@@@\seone@@@,@@@\ldots@@@,@@@\sen@@@)@@@}}

\com{\fnot}{{\rm@f@_0}}
\com{\fone}{{\rm@f@_1}}
\com{\ftwo}{{\rm@f@_2}}
\com{\fii}{{@f@_i}}
\com{\fn}{{@f@_n}}

\com{\cnot}{{\rm{\tt{}c}_0}}
\com{\cone}{{\rm{\tt{}c}_1}}
\com{\ctwo}{{\rm{\tt{}c}_2}}
\com{\ci}{{{\tt{}c}_i}}
\com{\cn}{{{\tt{}c}_n}}
\com{\anot}{{\rm@A@_0}}
\com{\aone}{{\rm@A@_1}}
\com{\atwo}{{\rm@A@_2}}
\com{\ai}{{@A@_i}}
\com{\aj}{{@A@_j}}
\com{\an}{{@A@_n}}
\com{\ta}{@A@}
\com{\te}{@E@}
\com{\tenot}{{@E@_{\it gl}}}
\com{\teone}{{\rm@E@_1}}
\com{\tei}{{@E@_i}}
\com{\ten}{{@E@_n}}
\com{\tnot}{{\rm@T@_0}}
\com{\tone}{{\rm@T@_1}}
\com{\ttwo}{{\rm@T@_2}}
\com{\tthree}{{\rm@T@_3}}
\com{\ti}{{@T@_i}}
\com{\tj}{{@T@_j}}
\com{\tn}{{@T@_n}}
\com{\ttt}{@T@}
\com{\unot}{{\rm@U@_0}}
\com{\uone}{{\rm@U@_1}}
\com{\utwo}{{\rm@U@_2}}
\com{\uthree}{{\rm@U@_3}}
\com{\ui}{{@U@_i}}
\com{\un}{{@U@_n}}
\com{\tunit}{@()@}
\com{\tint}{@Int@}
\com{\tfun}{\tone\ @->@\ \ttwo}
\com{\tfunp}{@(@\tone\ @->@\ \ttwo@)@}
\com{\thfun}{\tone\ \harrow\ \ttwo}
\com{\thfunp}{@(@\tone\ \harrow\ \ttwo@)@}
\com{\harrow}{@#>@}
\com{\tprod}{@(@\tone@,@\ldots@,@\tn@)@}
\com{\teprod}{@(@\teone@,@\ldots@,@\ten@)@}
\com{\tsum}{\cone\ \tone\ @+@\ \ldots\ @+@\ \cn\ \tn}
\com{\tdefn}{\aone\ @=@\ \tone@;@\ \ldots@;@\ \an\ @=@\ \tn}

@@

\com{\scone}{{\rm@@@c@@@_1}}
\com{\scn}{{@@@c@@@_n}}
\com{\saone}{{\rm@@@A@@@_1}}
\com{\san}{{@@@A@@@_n}}

\com{\stenot}{{@@@E@@@_{\it\subtypesubstyle{}gl}}}
\com{\sten}{{@@@E@@@_{\subtypesubstyle{}n}}}
\com{\stnot}{{@@@T@@@_{\rm\subtypesubstyle0}}}
\com{\stone}{{@@@T@@@_{\rm\subtypesubstyle1}}}
\com{\sttwo}{{@@@T@@@_{\rm\subtypesubstyle2}}}
\com{\sti}{{@@@T@@@_{\subtypesubstyle{}i}}}
\com{\stn}{{@@@T@@@_{\subtypesubstyle{}n}}}

\com{\sta}{{@@@A@@@}}
\com{\ste}{{@@@E@@@}}
\com{\stt}{{@@@T@@@}}
\com{\stu}{{@@@U@@@}}
\com{\std}{{@@@D@@@}}
\com{\stunit}{{@@@()@@@}}
\com{\stint}{{@@@Int@@@}}
\com{\stbool}{{@@@Bool@@@}}
\com{\stintlist}{{@@@IntList@@@}}
\com{\stfun}{{\stone@@@->@@@\sttwo}}
\com{\sthfun}{{\stone@@@#>@@@\sttwo}}
\com{\stprod}{{@@@(@@@\stone@@@,@@@\ldots@@@,@@@\stn@@@)@@@}}
\com{\steprod}{{@@@(@@@\steone@@@,@@@\ldots@@@,@@@\sten@@@)@@@}}
\com{\stsum}{{\scone\ \stone\ @@@+@@@\ \ldots\ @@@+@@@\ \scn\ \stn}}
\com{\stdefn}{{\saone\ @@@=@@@\ \stone@@@;@@@\ \ldots@@@;@@@\ \san\ @@@=@@@\ \stn}}

@@

\com{\ProjT}{{\it Proj}_{\stt}}
\com{\ProjU}{{\it Proj}_{\stu}}
\com{\SProjT}{{\it SProj}_{\stt}}
\com{\SProjU}{{\it SProj}_{\stu}}
\com{\SProjE}{{\it SProj}_{\ste}}
\com{\SAbsT}{{\it SAbs}_{\stt}}
\com{\FProjT}{{\it FProj}_{\stt}}
\com{\FProjU}{{\it FProj}_{\stu}}
\com{\FProjE}{{\it FProj}_{\ste}}
\com{\FAbsT}{{\it FAbs}_{\stt}}
\com{\FAbsFT}{{\it FAbsF_{\stt}}}
\com{\FTranT}{{\it FTran}_{\stt}}

\com{\Eprod}{@(@\Eone@,@\ldots@,@\En@)@}
\com{\Ei}{{@E@_i}}
\com{\Eone}{{\rm@E@_1}}
\com{\En}{{@E@_n}}

@@

\com{\oin}{{\rm1}\leq{}i\leq{}n}
\com{\ojn}{{\rm1}\leq{}j\leq{}n}
\com{\lubio}{\lub_{i\geq\rm0}}
\com{\glbio}{\glb_{i\geq\rm0}}
\com{\Lubio}{\Lub_{i\geq\rm0}}
\com{\Glbio}{\Glb_{i\geq\rm0}}
\com{\mkint}[1]{{\it{}mkint}^{#1}_{{\cal{}N}\br{@@@n@@@}}}
\rcom{\u}[1]{\acute{#1}}
\rcom{\d}[1]{\grave{#1}}
\com{\ud}[1]{\widehat{#1}}

@@

\com{\ve}{\vec{@e@}}
\com{\vx}{\vec{@x@}}
\com{\vy}{\vec{@y@}}
\com{\vT}{\vec{@T@}}
\com{\vF}{\vec{F}}
\com{\vv}{\vec{v}}

@@

\com{\fxge}{\mbox{@f x = @$\cdots$@(g e)@$\cdots$}}
\com{\C}{\mbox{$Dom$}}
\com{\Cn}{\mbox{$\C^n$}}
\com{\Cm}{\mbox{$\C^m$}}
\com{\Cs}{\mbox{$\C_s$}}
\com{\Csn}{\mbox{$\C_s^n$}}
\com{\Csm}{\mbox{$\C_s^m$}}

@@

\begin{abstract}
  {\it First-order\/} projection-based strictness analysis (generally,
  \emph{demand analysis}) was first introduced by Wadler and Hughes in
  1987~\cite{WH87}.  This development was restricted to first-order,
  monomorphic languages and provided no proof of soundness.  Subsequent
  developments extended the theory to other analyses (most notably
  binding-time analysis), approaches to higher order, and polymorphism.  Yet
  other developments were based on generalization of projections, specifically
  partial equivalence relations and partial projections, the latter notably
  simplifying a difficulty with projection-based strictness analysis dating
  back to Wadler and Hughes.  However, to date no development has combined the
  desirable attributes of all of these approaches.  We present the first
  practical, higher-order, polymorphic,
  partial-projection/partial-equivalence-relation based demand analysis,
  together with a proof of satisfaction of the fundamental safety condition
  for such analyses.  Such an analysis can be used both to improve serial code
  generation as well as enable safe automatic parallelization for lazy or
  non-strict functional languages.

% has proven genuinely useful in partial evaluation
% \cite{Lau91a}.  There have been three notable generalisations
% of projection-based analysis to higher order.  The first lacked a
% formal basis \cite{Mog89}; the second used structures strictly more
% general than projections, namely {\it partial equivalence relations\/}
% (PERs) \cite{HS91}; the third involved a complex construction that
% gave rise to impractically large abstract domains \cite{Dav93b}.  This
% paper presents a technique free of these shortcomings: it is simple,
% entirely projection-based, satisfies a formal correctness condition,
% and gives rise to reasonably small abstract domains.  Though the
% technique is cast in terms of projections, there is also an
% interpretation in terms of PERs.  The principal limitation of the
% technique is the restriction to {\it monomorphic\/} typing.
\end{abstract}

%\keywords{Binding-time analysis, abstract interpretation, projections, PERs.}

@@

We take as given that demand analysis, aka strictness analysis, is essential
for good serial code generation, and for identifying where parallelism may be
safely introduced.  An example of the former is the current ghc Haskell
compiler, where strictness analysis is used to unbox values and thereby
registerize code.  A potential application of the latter is in the context of
the APPFL parallelizing, which assumes strict semantics by default for the
purpose of exploring automatic parallelization techniques, but in the presence
of non-strict semantics could use such an analysis to enable such
parallelization.  Numerous such analysis techniques have been proposed and
implemented; we greatly narrow the field of discussion by restricting
attention to those for which there is a formally stated notion of correctness
that the technique has been proven to satisfy.

Analysis techniques can usually be identified as being based on either a
non-standard denotational semantics or a non-standard typing.  Examples in the
latter category include those of Gomard \cite{Go92}, Jensen \cite{Jen92}, the
Nielsons \cite{NN88}, Schmidt \cite{Sch88}, and Henglein and Mossin
\cite{HM94}.  Our focus is on those techniques based on non-standard
interpretation, in particular, those using projections, partial projections, or
partial equivalence relations (PERs) as the basic abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Wadler and Hughes first presented the idea of
projection-based program analysis in the context of

Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$
function $\BOT$, the least projection, which specifies that values are
entirely dynamic; and projections $\FST$ and $\SND$ on product
domains, defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\hspace{1.0in} \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs,
respectively.  The nominal goal of analysis is, given function $f$
denoted by some programming-language expression, and projection $\D$
encoding the staticness of the argument of $f$, to determine $\G$
satisfying the {\it safety condition\/} $\G\o\f\we\f\o\D$.  For
example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition, in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.

@@

For all closed expressions @e@ the abstract value
$\Ee{\Ps}{@e@}~[\,]$ of @e@ is of the form $(\ID,\K)$, showing
that closed expressions are always entirely static.  For expressions of
function type the abstract forward value $\K$ is a function from
abstract arguments of @e@ to abstract results.

>>>>

%% Kluwer Small Journal Article Sample File
%% June 20, 1994

%%   Use one of these two commands: 
\documentstyle{smjrnl} % For Computer Modern Fonts, or
%\documentstyle[smjfonts]{smjrnl} % for PostScript fonts

\begin{document}
\begin{article}

%%%%% To be entered at Kluwers: =====>>

\journame{Lisp and Symbolic Computation}
\volnumber{?}
\issuenumber{?}
\issuemonth{?}
\volyear{1995}

%% Do not delete either of the following two commands.
%% Please supply facing curly brackets for the part you
%% are not using for this article.
%\received{May 1, 1991}\revised{}

\authorrunninghead{K. Davis}
\titlerunninghead{PERs from Projections for Binding-time Analysis}

@@

%% <<== End of commands to be entered at Kluwers

%%  Authors, start here ==>>

@@

\com{\data}{{\it data}_{\subtypespace\stt}}
\com{\for}{{\it for}_{\subtypespace\stt}}
\com{\fac}{{\it fac}_{\subtypespace\stt}}
\com{\unfac}{{\it unfac}_{\subtypespace\stt}}

@@

\com{\tmn}{\theta_{mn}}
\com{\tnm}{\theta_{nm}}
\com{\tmy}{\theta_{m\infty}}
\com{\tyy}{\theta_{\infty\infty}}
\com{\tiy}{\theta_{i\infty}}
\com{\tym}{\theta_{\infty{}m}}
\com{\tyz}{\theta_{\infty{\rm0}}}
\com{\tyo}{\theta_{\infty{\rm1}}}
\com{\tyi}{\theta_{\infty{}i}}
\com{\tyip}{\theta_{\infty{}i+{\rm1}}}
\com{\th}{\theta}

@@

\com{\A}{\alpha}
\com{\lA}{\lift{\A}}
\com{\sA}{\slift{\A}}
\com{\B}{\beta}
\com{\lB}{\lift{\B}}
%\com{\sB}{\slift{\B}}
\com{\G}{\gamma}
\com{\lG}{\lift{\G}}
%\com{\sG}{\slift{\G}}
\com{\D}{\delta}
\com{\lD}{\lift{\D}}
\com{\sD}{\slift{\D}}
\com{\BOT}{{\it BOT}}
\com{\lBOT}{\lift{\BOT}}
\com{\sBOT}{\slift{\BOT}}
\com{\ID}{{\it ID}}
\com{\LAM}{{\it LAM}}
\com{\lID}{\lift{\ID}}
\com{\sID}{\slift{\ID}}
\com{\ABS}{{\it ABS}}
\com{\STR}{{\it STR}}
\com{\Ab}{\alpha_{\subtypespace{\makebox{\subtypesize\tt b}} }}
\com{\Ax}{\alpha_{\subtypespace{\makebox{\subtypesize\tt x}} }}
\com{\Ay}{\alpha_{\subtypespace{\makebox{\subtypesize\tt y}} }}
\com{\Az}{\alpha_{\subtypespace{\makebox{\subtypesize\tt z}} }}
\com{\Axs}{\alpha_{\subtypespace{\makebox{\subtypesize\tt xs}} }}
\com{\Ays}{\alpha_{\subtypespace{\makebox{\subtypesize\tt ys}} }}
\com{\Azs}{\alpha_{\subtypespace{\makebox{\subtypesize\tt zs}} }}

@@

\com{\oy}{\overline{\mbox{\tt y}}}

\com{\xnot}{{\rm\mbox{\tt x}_0}}
\com{\xone}{{\rm\mbox{\tt x}_1}}
\com{\xtwo}{{\rm\mbox{\tt x}_2}}
\com{\xthree}{{\rm\mbox{\tt x}_3}}
\com{\xxi}{{\mbox{\tt x}_i}}
\com{\xn}{{\mbox{\tt x}_n}}

\com{\ee}{\mbox{\tt e}}
\com{\ex}{\mbox{\tt x}}
\com{\evar}{\mbox{\tt x}}
\com{\enum}{\mbox{\tt n}_i}
\com{\esum}{\eone\ \mbox{\tt +}\ \etwo}
\com{\elam}{\mbox{\tt {\char'134}x.e}}
\com{\ehlam}{\mbox{\tt {\char'134}{\char'43}x.e}}
\com{\elamt}{\mbox{\tt {\char'134}x:T.e}}
\com{\ehlamt}{\mbox{\tt {\char'134}{\char'43}x:T.e}}
\com{\eapp}{\eone~\etwo}
\com{\ehapp}{\mbox{\tt app{\char'43}\ }\eone~\etwo}
\com{\efapp}{\mbox{\tt f\ e}}
\com{\eunit}{\mbox{\tt ()}}
\com{\etup}{\mbox{\tt (}\eone\mbox{\tt ,}\ldots\mbox{\tt ,}\en\mbox{\tt )}}
\com{\esel}{\mbox{\tt let\ (}\xone\mbox{\tt ,}\ldots\mbox{\tt ,}\xn\mbox{\tt )\ =}\ \enot\ \mbox{\tt in}\ \eone}
\com{\econ}{\mbox{\tt c}_i\ \mbox{\tt e}}
\com{\ecase}{\mbox{\tt case}\ \enot\ \mbox{\tt of\ }\cone\ \xone\ \mbox{\tt ->}\ \eone\mbox{\tt ;}\ \ldots\mbox{\tt ;}\ \cn\ \xn\ \mbox{\tt ->}\ \en}
\com{\efix}{\mbox{\tt fix\ e}}
\com{\ehfix}{\mbox{\tt fix{\char'43}\ e}}
\com{\estr}{\mbox{\tt seq}~\eone~\etwo}

\com{\enot}{{\rm\mbox{\tt e}_0}}
\com{\eone}{{\rm\mbox{\tt e}_1}}
\com{\etwo}{{\rm\mbox{\tt e}_2}}
\com{\ei}{{\mbox{\tt e}_i}}
\com{\en}{{\mbox{\tt e}_n}}

\com{\senot}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle0}}}
\com{\seone}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle1}}}
\com{\setwo}{{\rm{\makebox{\subtypesize\tt e}} _{\subtypesubstyle2}}}
\com{\sei}{{{\makebox{\subtypesize\tt e}} _{\subtypesubstyle{}i}}}
\com{\sen}{{{\makebox{\subtypesize\tt e}} _{\subtypesubstyle{}n}}}
\com{\setup}{{{\makebox{\subtypesize\tt (}} \seone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \sen{\makebox{\subtypesize\tt )}} }}

\com{\fnot}{{\rm\mbox{\tt f}_0}}
\com{\fone}{{\rm\mbox{\tt f}_1}}
\com{\ftwo}{{\rm\mbox{\tt f}_2}}
\com{\fii}{{\mbox{\tt f}_i}}
\com{\fn}{{\mbox{\tt f}_n}}

\com{\cnot}{{\rm{\tt{}c}_0}}
\com{\cone}{{\rm{\tt{}c}_1}}
\com{\ctwo}{{\rm{\tt{}c}_2}}
\com{\ci}{{{\tt{}c}_i}}
\com{\cn}{{{\tt{}c}_n}}
\com{\anot}{{\rm\mbox{\tt A}_0}}
\com{\aone}{{\rm\mbox{\tt A}_1}}
\com{\atwo}{{\rm\mbox{\tt A}_2}}
\com{\ai}{{\mbox{\tt A}_i}}
\com{\aj}{{\mbox{\tt A}_j}}
\com{\an}{{\mbox{\tt A}_n}}
\com{\ta}{\mbox{\tt A}}
\com{\te}{\mbox{\tt E}}
\com{\tenot}{{\mbox{\tt E}_{\it gl}}}
\com{\teone}{{\rm\mbox{\tt E}_1}}
\com{\tei}{{\mbox{\tt E}_i}}
\com{\ten}{{\mbox{\tt E}_n}}
\com{\tnot}{{\rm\mbox{\tt T}_0}}
\com{\tone}{{\rm\mbox{\tt T}_1}}
\com{\ttwo}{{\rm\mbox{\tt T}_2}}
\com{\tthree}{{\rm\mbox{\tt T}_3}}
\com{\ti}{{\mbox{\tt T}_i}}
\com{\tj}{{\mbox{\tt T}_j}}
\com{\tn}{{\mbox{\tt T}_n}}
\com{\ttt}{\mbox{\tt T}}
\com{\unot}{{\rm\mbox{\tt U}_0}}
\com{\uone}{{\rm\mbox{\tt U}_1}}
\com{\utwo}{{\rm\mbox{\tt U}_2}}
\com{\uthree}{{\rm\mbox{\tt U}_3}}
\com{\ui}{{\mbox{\tt U}_i}}
\com{\un}{{\mbox{\tt U}_n}}
\com{\tunit}{\mbox{\tt ()}}
\com{\tint}{\mbox{\tt Int}}
\com{\tfun}{\tone\ \mbox{\tt ->}\ \ttwo}
\com{\tfunp}{\mbox{\tt (}\tone\ \mbox{\tt ->}\ \ttwo\mbox{\tt )}}
\com{\thfun}{\tone\ \harrow\ \ttwo}
\com{\thfunp}{\mbox{\tt (}\tone\ \harrow\ \ttwo\mbox{\tt )}}
\com{\harrow}{\mbox{\tt {\char'43}>}}
\com{\tprod}{\mbox{\tt (}\tone\mbox{\tt ,}\ldots\mbox{\tt ,}\tn\mbox{\tt )}}
\com{\teprod}{\mbox{\tt (}\teone\mbox{\tt ,}\ldots\mbox{\tt ,}\ten\mbox{\tt )}}
\com{\tsum}{\cone\ \tone\ \mbox{\tt +}\ \ldots\ \mbox{\tt +}\ \cn\ \tn}
\com{\tdefn}{\aone\ \mbox{\tt =}\ \tone\mbox{\tt ;}\ \ldots\mbox{\tt ;}\ \an\ \mbox{\tt =}\ \tn}

@@

\com{\scone}{{\rm{\makebox{\subtypesize\tt c}} _1}}
\com{\scn}{{{\makebox{\subtypesize\tt c}} _n}}
\com{\saone}{{\rm{\makebox{\subtypesize\tt A}} _1}}
\com{\san}{{{\makebox{\subtypesize\tt A}} _n}}

\com{\stenot}{{{\makebox{\subtypesize\tt E}} _{\it\subtypesubstyle{}gl}}}
\com{\sten}{{{\makebox{\subtypesize\tt E}} _{\subtypesubstyle{}n}}}
\com{\stnot}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle0}}}
\com{\stone}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle1}}}
\com{\sttwo}{{{\makebox{\subtypesize\tt T}} _{\rm\subtypesubstyle2}}}
\com{\sti}{{{\makebox{\subtypesize\tt T}} _{\subtypesubstyle{}i}}}
\com{\stn}{{{\makebox{\subtypesize\tt T}} _{\subtypesubstyle{}n}}}

\com{\sta}{{{\makebox{\subtypesize\tt A}} }}
\com{\ste}{{{\makebox{\subtypesize\tt E}} }}
\com{\stt}{{{\makebox{\subtypesize\tt T}} }}
\com{\stu}{{{\makebox{\subtypesize\tt U}} }}
\com{\std}{{{\makebox{\subtypesize\tt D}} }}
\com{\stunit}{{{\makebox{\subtypesize\tt ()}} }}
\com{\stint}{{{\makebox{\subtypesize\tt Int}} }}
\com{\stbool}{{{\makebox{\subtypesize\tt Bool}} }}
\com{\stintlist}{{{\makebox{\subtypesize\tt IntList}} }}
\com{\stfun}{{\stone{\makebox{\subtypesize\tt ->}} \sttwo}}
\com{\sthfun}{{\stone{\makebox{\subtypesize\tt {\char'43}>}} \sttwo}}
\com{\stprod}{{{\makebox{\subtypesize\tt (}} \stone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \stn{\makebox{\subtypesize\tt )}} }}
\com{\steprod}{{{\makebox{\subtypesize\tt (}} \steone{\makebox{\subtypesize\tt ,}} \ldots{\makebox{\subtypesize\tt ,}} \sten{\makebox{\subtypesize\tt )}} }}
\com{\stsum}{{\scone\ \stone\ {\makebox{\subtypesize\tt +}} \ \ldots\ {\makebox{\subtypesize\tt +}} \ \scn\ \stn}}
\com{\stdefn}{{\saone\ {\makebox{\subtypesize\tt =}} \ \stone{\makebox{\subtypesize\tt ;}} \ \ldots{\makebox{\subtypesize\tt ;}} \ \san\ {\makebox{\subtypesize\tt =}} \ \stn}}

@@

\com{\ProjT}{{\it Proj}_{\subtypespace\stt}}
\com{\ProjU}{{\it Proj}_{\subtypespace\stu}}
\com{\SProjT}{{\it SProj}_{\subtypespace\stt}}
\com{\SProjU}{{\it SProj}_{\subtypespace\stu}}
\com{\SProjE}{{\it SProj}_{\subtypespace\ste}}
\com{\SAbsT}{{\it SAbs}_{\subtypespace\stt}}
\com{\FProjT}{{\it FProj}_{\subtypespace\stt}}
\com{\FProjU}{{\it FProj}_{\subtypespace\stu}}
\com{\FProjE}{{\it FProj}_{\subtypespace\ste}}
\com{\FAbsT}{{\it FAbs}_{\subtypespace\stt}}
\com{\FAbsFT}{{\it FAbsF_{\subtypespace\stt}}}
\com{\FTranT}{{\it FTran}_{\subtypespace\stt}}

\com{\Eprod}{\mbox{\tt (}\Eone\mbox{\tt ,}\ldots\mbox{\tt ,}\En\mbox{\tt )}}
\com{\Ei}{{\mbox{\tt E}_i}}
\com{\Eone}{{\rm\mbox{\tt E}_1}}
\com{\En}{{\mbox{\tt E}_n}}

@@

\com{\oin}{{\rm1}\leq{}i\leq{}n}
\com{\ojn}{{\rm1}\leq{}j\leq{}n}
\com{\lubio}{\lub_{i\geq\rm0}}
\com{\glbio}{\glb_{i\geq\rm0}}
\com{\Lubio}{\Lub_{i\geq\rm0}}
\com{\Glbio}{\Glb_{i\geq\rm0}}
\com{\mkint}[1]{{\it{}mkint}^{#1}_{{\cal{}N}\br{{\makebox{\subtypesize\tt n}} }}}
\rcom{\u}[1]{\acute{#1}}
\rcom{\d}[1]{\grave{#1}}
\com{\ud}[1]{\widehat{#1}}

@@

\com{\ve}{\vec{\mbox{\tt e}}}
\com{\vx}{\vec{\mbox{\tt x}}}
\com{\vy}{\vec{\mbox{\tt y}}}
\com{\vT}{\vec{\mbox{\tt T}}}
\com{\vF}{\vec{F}}
\com{\vv}{\vec{v}}

@@

\com{\fxge}{\mbox{\mbox{\tt f\ x\ =\ }$\cdots$\mbox{\tt (g\ e)}$\cdots$}}
\com{\C}{\mbox{$Dom$}}
\com{\Cn}{\mbox{$\C^n$}}
\com{\Cm}{\mbox{$\C^m$}}
\com{\Cs}{\mbox{$\C_s$}}
\com{\Csn}{\mbox{$\C_s^n$}}
\com{\Csm}{\mbox{$\C_s^m$}}

@@

\title{PERs from Projections for Binding-time Analysis}

\authors{Kei Davis}
\email{kei@crl.nmsu.edu}

\affil{Center for Non-Linear Studies \\
Mail Stop B258 \\
Los Alamos National Laboratory \\
Los Alamos, NM  87545, U.S.A.}

%\editor{Dennis Kibler}

\abstract{{\it First-order\/} projection-based binding-time analysis
has proven genuinely useful in partial evaluation
\cite{Lau91a}.  There have been three notable generalisations
of projection-based analysis to higher order.  The first lacked a
formal basis \cite{Mog89}; the second used structures strictly more
general than projections, namely {\it partial equivalence relations\/}
(PERs) \cite{HS91}; the third involved a complex construction that
gave rise to impractically large abstract domains \cite{Dav93b}.  This
paper presents a technique free of these shortcomings: it is simple,
entirely projection-based, satisfies a formal correctness condition,
and gives rise to reasonably small abstract domains.  Though the
technique is cast in terms of projections, there is also an
interpretation in terms of PERs.  The principal limitation of the
technique is the restriction to {\it monomorphic\/} typing.}

\keywords{Binding-time analysis, abstract interpretation, projections,
PERs.}

@@

We take as given that binding-time analysis is essential for good
partial evaluation, and we do not address the issue of annotating
programs according to the results of analysis.  Numerous binding-time
analysis techniques have been proposed and implemented; we greatly
narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the
technique has been proven to satisfy.

Analysis techniques can usually be identified as being based on either
a non-standard denotational semantics or a non-standard typing.
Examples in the latter category include those of Gomard \cite{Go92},
Jensen \cite{Jen92}, the Nielsons \cite{NN88}, Schmidt \cite{Sch88},
and Henglein and Mossin \cite{HM94}.  Our focus is on those techniques
based on non-standard interpretation, in particular, those using
projections or partial equivalence relations (PERs) as the basic
abstract values.

A domain {\it projection\/} is a continuous idempotent function that
approximates the identity.  Launchbury \cite{Lau88} hit upon the idea
of using projections to encode degrees of staticness of data.  The
basic idea is that a projection maps to $\bot$ that part of a data
structure that is dynamic (possibly not determined), and acts as the
identity on that part which is static (definitely determined).
Examples are the identity $\ID$, the greatest projection, which
specifies that values are entirely static; the constant $\bot$
function $\BOT$, the least projection, which specifies that values are
entirely dynamic; and projections $\FST$ and $\SND$ on product
domains, defined by
\[
\FST\ (x,y) = (x,\bot)\ ,\hspace{1.0in} \SND\ (x,y) = (\bot,y)\ ,
\]
specifying staticness in the first and second components of pairs,
respectively.  The nominal goal of analysis is, given function $f$
denoted by some programming-language expression, and projection $\D$
encoding the staticness of the argument of $f$, to determine $\G$
satisfying the {\it safety condition\/} $\G\o\f\we\f\o\D$.  For
example, taking $\G$ to be $\SND$ satisfies
$\G\o\itt{swap}\we\itt{swap}\o\FST$ for $\itt{swap}$ defined by
$\itt{swap}~(x,y)=(y,x)$.  Taking $\G$ to be $\BOT$ always satisfies
the safety condition but tells nothing; greater $\G$ is more
informative.  Launchbury \cite{Lau91a} showed that this safety
condition, in a sense which he formalises, is equivalent
to the correctness condition for binding-time analysis in the general
framework of Jones \cite{Jon88}.  Using projection-based analysis,
Launchbury implemented both monomorphic and polymorphic versions of a
partial evaluator for a first-order language.

@@

For all closed expressions \mbox{\tt e} the abstract value
$\Ee{\Ps}{\mbox{\tt e}}~[\,]$ of \mbox{\tt e} is of the form $(\ID,\K)$, showing
that closed expressions are always entirely static.  For expressions of
function type the abstract forward value $\K$ is a function from
abstract arguments of \mbox{\tt e} to abstract results.
